A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-02-14 jacobian_calc_CMAME.tex
I.3 addressed, I.2 partially addressed

\note[Michael]{re: I.2, John I think you better handle expanding this section,
I was going to essentially copy silling:psa.}\change[Michael]{Briefly, peridynamics}Peridynamics \cite{silling2000ret,silling:psa,silling2010peridynamic}
is a nonlocal reformulation of the partial-differential equations that provide
the statement of momentum balance in classical continuum mechanics. Its primary
goal is to avoid the use of spatial derivatives in the balance equations or
constitutive laws such that discontinuous displacements, i.e. cracks, are
mathematically consistent with the governing equations. It has demonstrated
great promise in modeling problems with pervasive failure
\cite{littlewood2010}, interesting phenomenon such as dynamic crack branching
\cite{ha2010sod}, and is being extended to multiphyics phenomena
\cite{bobaru2011peridynamic,katiyar2013} and the problem of multi-scale
coupling to molecular simulations \cite{seleson2009peridynamics}. The related mathematical
definitions and parts of peridynamic theory especially relevant to the present
work are explained in the sequel.

@@

For the finite-difference type methods in \emph{Peridigm}, the perturbation method is employed
by probing the deformed configuration one degree of freedom at a time and so allowing the
sensitivity of the force density state wrt the deformation state for all nodes to be computed.
}
\subsubsection{Implementing the complex-step method in Peridigm} 
%
The CS method was implemented in \emph{Perdigm} in a manner that follows
closely the implementation of the FD technique.  However, because of
\emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the
\emph{Epetra} package, special steps had to be followed to allow the use of
complex number data types.  This is because \emph{Epetra} vectors, which can be
thought of as distributed memory parallel cousins to standard C++ STL vectors,
are hard coded to be of type {\tt double} and can not be simply be declared as
having a complex data type. Therefore, program methods that were used in the
course of applying the FD method were copied, renamed and re-written to follow
the CS formula. This involved changing the data type of intermediate variables
which were locally scoped, and dynamically casting persistent memory variables
to complex data types where necessary. No attempt was made to ``tune'' the CS
code in any way for performance. The algorithm  was implemented in such as way
that it was in one-to-one correspondence with the FD and CD methods; however,
it must be noted that the overhead associated with the dynamic casts in the CS
method likely hurts its overall performance and this issue could be alleviated
by using the next generation of templated \emph{TPetra} vectors available in
\emph{Trilinos} that are capable of taking an explicit instantiation of any
data type including {\tt complex}. The version of \emph{Peridigm} modified to
implement CS can be found in the paper repository.

\emph{Peridigm} was primarily written by computational scientists; therefore,
many advanced techniques in C++ such as multiple virtual inheritance, pointer
arithmetic, and distributed data structures were used in the basic code and by
necessity in the modified code.  To see CS and other methods in a more narrow
context, it is advised that readers also take a look at the simple examples
provided on the corresponding author's website.

@@

Another series of four test problems were simulated, however this time the
number of compute cores used to solve the problem was increased from test to
test while the discretization level was held constant at 1 million
computational nodes (3 million degrees of freedom). This allowed for sufficient
computational nodes per core even at the highest level of parallelization such
that message passing computation did not overwhelm the simulation.   The
specific parameters of these tests can be found in \emph{Perdigm} input {\tt
xml} files also included in the paper repository.

@@

The test problems were run on the \emph{Stampede} HPC cluster computer housed
at \emph{TACC} (Texas Advanced Computing Center) at The University of Texas at
Austin. The single core test series was run using the {\tt normal} queue
compute nodes, while the parallel test series was run using one to four {\tt
large memory} nodes each having one terabyte of local RAM. This large amount of
memory was required for reasons explained in Section~\ref{JGAM}.

@@

The CS method was shown to be highly accurate in calculating tangent-stiffness
matrices by the standards of this study, and it is slower than other methods
for a naive implementation. In the context of future use of CS in
\emph{Peridigm}, CS holds the advantage of relying on byte-copyable data types,
while AD requires a complicated serialization and deserialization in order to
function as byte-copyable. This is fine for copying in between MPI processes,
but deserialization on an accelerator or GPU is very difficult, as this would
require compiling a version of the AD library being used that were compatible
with the accelerator or GPU being used. What this means is that if
\emph{Peridigm} or any simulation software were to be reformulated to take
advantage of accelerators or GPUs, the flexibility of complex-step owed to its
simple implementation may make it more viable for that application given its
high accuracy. Additional performance gains may be achieved through the use of
the templated \emph{TPetra} package in \emph{Trilinos} which would allow the
use of complex data types natively, and alleviate the expense of the
dynamic-casts that were utilized in this study.  As a final comment, it should
be noted that from a code development perspective the CS method may be the
easiest of the methods to implement; It only requires one functional
evaluation, there is no heuristic choice of probe distance, and there is no
dependence on an external AD library.  From this perspective, the method may
still be an attractive method for TS matrix evaluation.

>>>>

\note[Michael]{re: I.2, John I think you would better handle expanding this section,
I was going to essentially copy silling:psa.}\change[Michael]{Briefly, peridynamics}Peridynamics \cite{silling2000ret,silling:psa,silling2010peridynamic}
is a nonlocal reformulation of the partial-differential equations that provide
the statement of momentum balance in classical continuum mechanics. Its primary
goal is to avoid the use of spatial derivatives in the balance equations or
constitutive laws such that discontinuous displacements, i.e. cracks, are
mathematically consistent with the governing equations. It has demonstrated
great promise in modeling problems with pervasive failure
\cite{littlewood2010}, interesting phenomenon such as dynamic crack branching
\cite{ha2010sod}, and is being extended to multiphyics phenomena
\cite{bobaru2011peridynamic,katiyar2013} and the problem of multi-scale
coupling to molecular simulations \cite{seleson2009peridynamics}. The related mathematical
definitions and parts of peridynamic theory especially relevant to the present
work are explained in the sequel.

@@

For the finite-difference type methods in \emph{Peridigm}, the perturbation
method is employed by probing the deformed configuration one degree of freedom
at a time and so allowing the sensitivity of the force density state wrt the
deformation state for all nodes to be computed.  }

\subsubsection{Implementing the complex-step method in Peridigm} 
%
The CS method was implemented in \emph{Perdigm} in a manner that follows
closely the implementation of the FD technique.  However, because of
\emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the
\emph{Epetra} package, special steps had to be followed to allow the use of
complex number data types.  This is because \emph{Epetra} vectors, which can be
thought of as distributed memory parallel cousins to standard C++
\change[Michael]{STL}{Standard Template Library} vectors, are hard coded to be
of type {\tt double} and can not be simply be declared as having a complex data
type. \note[Michael]{re: I.3}\add[Michael]{For the reader's convenience,
    \emph{Epetra} vectors and STL vectors are what are known as 'container
    classes', whose purpose in object oriented computer programming is to
    automate the creation, accessing, modification and destruction of the
    underlying contained data. Templated container classes allow for user
    choice in what properties that underlying data exihibits.  For example, in
    this work, complex arithmetic cannot be performed on the data contained
    within an \emph{Epetra} vector because that data, of type {\tt double} does
    not have the property of following the rules of complex arithmetic. The
    reason the non-templated \emph{Epetra} vector could not be dispensed with
    for the purposes of this work is because the \emph{Epetra} vector was
tightly integrated with \emph{Peridigm's} parallel communication strategy and
linear algebra code.} Program methods that were used in the course of applying
the FD method were copied, renamed and re-written to follow the CS formula.
Because of restrictions imposed by the container class being used, this
involved changing the data type of intermediate variables which were locally
scoped, and dynamically casting persistent memory variables to complex data
types where necessary. \note[Michael]{re: I.3}\add[Michael]{For the reader's
convenience, 'locally scoped' in the C++ programming language refers to
data which is created within a context more specific than the context of
the entire program, or 'translation unit'.  Most data created this way is
destroyed when that specific context is left. The topic of contexts in C++
is very broad, so for the purposes of this work context refers to a program
method or function that begins, concludes and yields a result, also
destroying locally scoped data. Dynamic casting for the purposes of this
work refers to copying the value of data contained within an \emph{Epetra}
vector into locally scoped data that has the property of following the rules of
complex arithmetic, where a mapping called 'type casting' allows the
translation of data between different types.} No attempt was made to ``tune''
the CS code in any way for performance. The algorithm  was implemented in such
as way that it was in one-to-one correspondence with the FD and CD methods;
however, it must be noted that the overhead associated with the dynamic casts
in the CS method likely hurts its overall performance and this issue could be
alleviated by using the next generation of templated \emph{TPetra} vectors
available in \emph{Trilinos} that are capable of taking an explicit
instantiation of any data type including {\tt complex}. \note[Michael]{re:
I.3}\add[Michael]{For the purposes of this work, an explicit instantiation
refers to requesting the creation of an object, or instance, of a templated
class or container class that holds data which has the property of
following desired mathematic or logic rules.} The version of \emph{Peridigm}
modified to implement CS can be found in the paper repository.

\emph{Peridigm} was primarily written by computational scientists; therefore,
many advanced techniques in C++ such as multiple virtual inheritance, pointer
arithmetic, and distributed data structures, \note[Michael]{re:
I.3}\add[Michael]{for example \emph{Epetra} vectors, were used in the basic
    code and by necessity in the modified code. These concepts are not vital to
understanding the complex step method iteself, but may be necessary to understand complex
step in \emph{Peridigm}.} To see CS and other methods in a more narrow context,
it is advised that readers also take a look at the simple examples provided on
the corresponding author's website.

@@

Another series of four test problems were simulated, however this time the
number of compute cores used to solve the problem was increased from test to
test while the discretization level was held constant at 1 million
computational nodes (3 million degrees of freedom). This allowed for sufficient
computational nodes per core even at the highest level of parallelization such
that message passing computation did not overwhelm the simulation.  The
specific parameters of these tests can be found in \emph{Perdigm} input {\tt
xml} files also included in the paper repository. \note[Michael]{re:
I.3}\add[Michael]{For the purposes of this work message passing is what
synchronizes multiple computers that each are working on a separate section
of the problem domain. This operation involves the time-consuming use of
the data network. To reduce the proportion of simulation time spent on
message passing, good practice is to maximize the amount of information
contained within a subdomain solely managed by a single computer, relative
to the amount of information shared accross multiple subdomains and
therefor multiple computers. A useful experiment to illustrate this idea
concretely would be to implement a code to solve a 2D Laplace equation for
simulating a simple heat condution problem on two processors. Use the
textbook finite difference method, which in this case does not refer just
to the differentiation technique.  First decompose the domain into two
contiguous chunks, having one shared border, and distribute these chunks,
or subdomains to the two processors. Perfom the simulation a multitude of
times to calculate an average runtime.  Next implement the solution a
different way, instead partitioning the problem domain into many narrow,
alternating strips, and distribute those to again two processors, where adjacent
strips in the graph of the domain should be sent to different processors. This way
of decomposing the domain will cause much overlap between the subdomains
managed by the two processors and much time will be spent synchronzing the
subdomain overlap or 'ghost nodes' between the processors. Run the modified
simulation a multitude of times and take the average runtime. The larger
the unpartitioned domain, the worse the discrepancy should be.}

@@

The test problems were run on the \emph{Stampede} HPC cluster computer housed
at \emph{TACC} (Texas Advanced Computing Center) at The University of Texas at
Austin. The single core test series was run using the {\tt normal} queue
compute nodes, while the parallel test series was run using one to four {\tt
large memory} nodes each having one terabyte of local RAM. This large amount of
memory was required for reasons explained in Section~\ref{JGAM}, second paragraph.

@@

The CS method was shown to be highly accurate in calculating tangent-stiffness
matrices by the standards of this study, and it is slower than other methods
for a naive implementation. In the context of future use of CS in
\emph{Peridigm}, CS holds the advantage of relying on byte-copyable data types,
while AD requires a complicated serialization and deserialization in order to
function as byte-copyable. This is fine for copying in between MPI processes,
but deserialization on an accelerator or GPU is very difficult, as this would
require compiling a version of the AD library being used that were compatible
with the accelerator or GPU being used. \note[Michael]{re:
I.3}\add[Michael]{For the reader's convenience, a collection of contiguously
stored data elements are said to be 'byte-copiable' when the the interval
in bytes of memory spanned by a data element in the collection is constant
throughout the collection. Suppose 'contiguously stored' means, stored in
memory with a minimum of unrelated infomation stored in between data
elements in the collection. Complicated objects, such as AD datatype
elements, do not have the same size in memory as other objects even of that
same class, for the reason that they themselves contain an uncertain
ammount of state data according to their design and work done in the
program. This means that a collection of such objects, even if they are
contiguously stored in memory, will not be alligned according to a constant
interval, such that elements within a collection cannot be located by
coordinates consisting of the location of the beginning of the first
element in the collection and an interval multiplied by a natural number.
This method of locating is integral to the serialized communication
implemented by MPI the libraries as well as CUDA, OpenCL, and others,
which allow only the transmission of collections of byte-copiable datatypes
(that is 'primitive datatypes'). A common method of bypassing this
limitation for communication purposes is to write the complicated data to a
'character buffer', and transmit the byte-copiable character buffer,
however this is useless unless the recipient device of the communication
can recover the original data and process it, which is impossible if the
recipient device only can work with a subset of the programming language
that the sending device must use, such as in a communication between a CPU
running C++ and a GPU running C (such as in CUDA programs), where the gap in
sophistication contains the features of the "complicated" datatype.} What this means is
that if \emph{Peridigm} or any simulation software were to be reformulated to
take advantage of accelerators or GPUs, the flexibilty of complex-step owed to
its simple implementation may make it more viable for that application given
its high accuracy. Additional performance gains may be achieved through the use
of the templated \emph{TPetra} package in \emph{Trilinos} which would allow the
use of complex data types natively, and alleviate the expense of the
dynamic-casts that were utilized in this study.  As a final comment, it should
be noted that from a code development perspective the CS method may be the
easiest of the methods to implement; It only requires one functional
evaluation, there is no heuristic choice of probe distance, and there is no
dependence on an external AD library.  From this perspective, the method may
still be an attractive method for TS matrix evaluation.
